Co-methylation

### 2014-10-08 Rmarkdown/comethylation.Rmd
Split co-methylation chapter.
Now split into “old stuff” (review of previous work) and “new stuff” (my work)

Co-methylation is the dependence of DNA methylation at different methylation loci within a sample. As such, it is a kind of spatial dependence where the "space" is the genome of the sample. In this chapter I describe two measures of co-methylation, namely, _within-fragment co-methylation_ and _correlation of $\beta$-values_. Using both these measures I explore how co-methylation is affected by the distance between methylation loci, the density of methylation loci and the genomic context of the methylation loci.

I begin by explaining various methods for estimating co-methylation and their associated challenges. I then use these methods to analyse more than fifty whole-genome bisulfite-sequencing libraries to study the variation of co-methylation within and between different cell types.

@@

\section{Previous work}\label{subsec:previous work}

I define co-methylation as the dependence of DNA methylation at different methylation loci within a sample. In other words, it is the dependence or correlation structure of DNA methylation data within a sample. This is slightly different to \cite{Schatz:2004cha} who defined it as, "the presence of methylation over a stretch of neighbouring CpG positions"; it is more in line with \cite{Eckhardt:2006gh} who defined it as, "the relationship between the degree of methylation over distance".

There have been several attempts to explore the relationship between methylation events at different loci within an individual. However, frequently the exact question being asked and the methodology have not been well-defined. Moreover, to the best of my knowledge, there is no software available that implements any of the proposed methods.

From a biological perspective, by studying co-methylation we hope to better understand the regulation of DNA methylation. From a statistical perspective, more powerful and efficient analysis methods can be developed if this correlation structure is better understood, accounted for and exploited.

An increased understanding of co-methylation could also lead to cheaper and more efficient assays of DNA methylation. For example, we could identify methylation loci whose methylation status is highly predictive of the methylation level of a surrounding region and design an assay to interrogate those highly predictive loci. This idea is analogous to the use of tag-SNPs on whole-genome genotyping microarrays (__CITE__).

\subsection{Correlations of aggregate measurements of methylation}
\label{sec:previous_cor_of_agg_methylation}

To date, most analyses of co-methylation have been based on analysing aggregate measurements of methylation, such as $\beta$-values, within individual samples \cite{Eckhardt:2006gh, Cokus:2008fc, Li:2010fb, Lyko:2010dr, Lacey:2013iy}. These analyses of within-sample correlations of aggregate methylation levels have an implicit assumption of stationarity along the genome (__TODO: Discuss with Terry__).

Here I review this previous work. I have been forced to make my "best guess" about some of the methods since they are often not sufficiently described in the original publication.

#### \cite{Eckhardt:2006gh}

\cite{Eckhardt:2006gh} is an important paper and one of the first to apply bisulfite-sequencing to a large number of regions and samples. Of particular relevance to a discussion of co-methylation, \cite{Eckhardt:2006gh} has been cited as evidence that "methylation levels are strongly spatially correlated" \cite{Hebestreit:2013ko} and that "proximal CpGs [have] similar methylation levels" \cite{Hansen:2011gu}. \cite{Eckhardt:2006gh} performed  PCR-amplicon bisulfite-sequencing of 2,524 amplicons on human chromosomes 6, 20 and 22.  PCR-amplicon bisulfite-sequencing is a labour-intensive assay and so it was a considerable effort to produce these data. Nonetheless, it is important to bear in mind that these data are nowhere near as comprehensive as data generated using modern-day whole-genome bisulfite-sequencing assays. Samples were provided by male and female donors across a wide variety of tissue types and ages, and were were generated to begin a catalogue of CpG methylation in various tissue types.

As part of their analysis, the authors analyse co-methylation, which they define as "the relationship between the degree of methylation over distance". They state that "we were able to establish a significant correlation for comethylation over short distances ($\leq 1,000$ bp), it deteriorated rapidly for distances $ >2,000$ bp" and reference Figure 3a in support of this claim. However, what is plotted in Figure 3a is not "correlation" in the usual sense of the word.

In Figure 3a, the authors plot the "percentage identical methylation" against "distance in bp", which ranges from $0$ to $20,000$ bp (__TODO: Check permission on re-priting this figure__). The average amplicon length is reported as $411$ bp with a standard deviation of $71$ bp. Therefore, it is not possible that the "percentage identical methylation" was obtained from methylation events co-occuring on the same amplicon, at least for those pairs separated by more than approximately $600$ bp. Nowhere in the original paper is it clearly described what the "percentage identical methylation" is nor how it is computed. An email discussion with the senior author, Stephan Beck, was unable to clarify the definition.

My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of methylation calls at different CpGs from (generally) different amplicons where the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$. I will refer to my best guess of "percentage identical methylation" as $PIM^{*}$, and now turn to some of its properties.

Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $z_{i, h}$ is sampled, that is, $Pr(z_{h, i} = 1) = p_{h}$. For two distinct CpGs sequenced on two (potentially) distinct amplicons, $z_{h, i}$ and  $z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$. Not only may the two CpGs be from distinct amplicons, but the amplicons themselves may be from two distinct samples with different genome-wide average methylation levels. This most general setting is described by $z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $z_{h, i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(z_{h, i}, z_{h', i'}) = c$. In this setting we can compute the theoretical $PIM^{*}$ as follows:

\begin{align*}
PIM^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
        &= Pr(z_{h, i} = 0, z_{h', i'} = 0) + Pr(z_{h, i} = 1, z_{h', i'} = 1) \\
        &= Pr(z_{h', i'} = 0 | z_{h, i} = 0) Pr(z_{h, i} = 0) +
           Pr(z_{h', i'} = 1 | z_{h, i} = 1) Pr(z_{h, i} = 1) \\
        &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
           \big [ (p_{h'} + c (1 - p_{h})) \big ] p_{h} \\
        &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
\end{align*}

This result follows by substituting $z_{h, i}$ and $p_{h}$ (resp. $z_{h', i'}$ and $p_{h'}$) for $X_{1}$ and $p_{1}$ (resp. $X_{2}$ and $p_{2}$) into the result derived in __TODO: Cite and link to appendix__.

__TODO: What is the general restriction on $c$ when $p_{h, i} \neq p_{h', i'}$?__

In general, $PIM^{*} \in [0, 1]$. However, there are a few important special cases that are worth discussing:

1. $PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $z_{h, i} \eqd Bernoulli(p_{h})$ and $z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(z_{h, i}, z_{h', i'}) = c = 0$.
2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c \neq 0$.
3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c = 0$.

As for $PIM^{*}$, $PIM^{*}_{inid} \in [0, 1]$. In contrast, it is easy to show by differentiation that $PIM^{*}_{iid} \geq 0.5$ and that $PIM^{*}_{did} \geq 0.5 (c + 1)$, which is $> 0.5$ (resp. $< 0.5$) for $c > 0$ (resp. $c < 0$). __TODO: Illustrate these special cases with a plot.__

To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{*}_{iid}$. If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp. However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible even supposing $c > 0$.

Of course, the above discussion pre-supposes that what is labelled "percentage identical methylation" in Figure 3a is mathematically equivalent to $PIM^{*}$. I believe this to be true but due to the aforementioned deficiencies of the method description it is not possible to verify this derivation.

Quite apart from what exactly is plotted in Figure 3a, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis. Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure 3a does not represent any one sample but a pool of samples.

#### \cite{Cokus:2008fc}\label{sec:Cokus}

\cite{Cokus:2008fc} published the BS-seq protocol for performing whole-genome bisulfite-sequencing. They demonstrated this technique by generating a genome-wide map of DNA methylation in wildtype  _Arabidopsis Thaliana_. They performed an autocorrelation analysis of the $\beta$-values for CG, CHG and CHH methylation[^arabidopsis]. The autocorrelation computes, for a set of distances ($0-5,000$ nt), the (Pearson) correlation coefficient of methylation levels for all pairs of methylation loci separated by each distance. To emphasise, I believe this analysis uses all pairs of methylation loci, that is, regardless of how many intervening methylation loci there are for each pair.

[^arabidopsis]: Many plant species, such as _A. thaliana_, have much higher levels of non-CG methylation than do mammals.

\cite{Cokus:2008fc} found that methylation levels are highly correlated, particularly for CG methylation levels, and that the correlation decays as a function of genomic distance between methylation loci. From this they concluded that the "significant correlation between methylated cytosines for distances up to 5,000 nucleotides or more [were] probably a reflection of regional foci of methylation throughout the genome and of large blocks of pericentromeric heterochromatin". They also identified correlations between different methylation contexts, e.g. CG vs. CHG, which "suggest[s] complex interactions between the different types of methylation".

The autocorrelation plots showed two periodicities, one of $\approx 167$ nucleotides and the other of $10$ nucleotides (CHH only). Both periodicities are visible in plots of the raw autocorrelations and are confirmed by a Fourier transform analysis of the autocorrelation.

[^CHH_only]: The $10$ nucleotide periodicity was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}

The $167$ nucleotide periodicity is "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al. hypothesise that this periodicity is due to histones dictating access to DNA by the DNA methyltransferases.

The $10$ nucleotide periodicity is equal to the length of one helical DNA turn. From this result Cokus et al. hypothesise that the DOMAINS REARRANGED METHYLASE 2 (_DRM2_), "the main enzyme controlling asymmetric methylation in _Arabidopsis_", contains two active sites that methylate sites 8-10 apart[^comethylation_def2], like its mammalian homologue, DNA methyltransferase 3 (_DNMT3_) (__TODO: If I want a picture explaining this then Figure 3b from \cite{Jurkowska:2011dt} might form the basis for my own illustration). However, in a later paper by the same group, this $10$ nucleotide periodicity was also observed for CG and CHG methylation, at least for nucleosome bound DNA \cite{Chodavarapu:2010iqa}.

[^comethylation_def2]: This is referred to as "co-methylation", here meaning the methylation of multiple loci in a single event, by \cite{Jurkowska:2011dt}. This article reviews a plausible biochemical model for how _DNMT3a_ could co-methylate two CpGs separated by $8-10$ nucleotides; this model includes both the scenario where both CpGs are on the same DNA strand and the scenario where the CpGs are on opposite strands.

The $10$ nucleotide CHH-only periodicity in the autocorrelation analysis of the $\beta$-values is also observed in a _within-fragment analysis_ (discussed in \ref{sec:previous_within-fragment_analyses})

#### \cite{Chodavarapu:2010iqa}

\cite{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (HSF1). The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq. The nucleosome start sites anchor the methylation observations to a common grid and is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes. A $10$ nucleotide periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}. Again, the observed periodicity is confirmed by analysing a Fourier transform of these plots.

In addition to the genome-wide analyses in both _Arabidopsis_ and human embryonic stem cell line (HSF1), the same method is used to examine promoters, genes, repeats, euchromatic regions and centromeric regions of _Arabidopsis_ and promoters, genes, repeats and CpG islands of the HSF1 sample, which confirmed the $10$ nucleotide periodicity. This led \cite{Chodavarapu:2010iqa} to propose an extended version of their earlier hypothesis \cite{Cokus:2008fc}: namely, "that nucleosomes are to some extent dictating access to the DNA and therefore setting the register of methylation __for all DNA methyltransferases__" (emphasis mine) and not just for the DRM2 methyltransferase.

#### \cite{Li:2010fb}

\cite{Li:2010fb} compute "the correlation of methylation level of any two nearby CpGs and the relationship between spatial distance (from one CpG to another) and strength of this correlation". The authors conclude that "co-methylation deteriorates over distance and becomes nearly undetectable at distances $> 1,000$ bp", citing their results as being consistent with those in \cite{Eckhardt:2006gh}. In addition to this genome-wide analysis, \cite{Li:2010fb} perform the same analysis for CpGs within 19 different genomic contexts such as upstream of a gene, untranslated region, coding DNA sequences and various repetitive elements. Somewhat surprisingly, they do not specifically compare CpG islands to non CpG islands, although some of these differences will be captured by the analysis of regions upstream of a gene, which include many CpG island promoters.

Unfortunately, just as in \cite{Eckhardt:2006gh}, \cite{Li:2010fb} never define exactly what is being computed in their analysis of co-methylation and no software is available. All the co-methylation results are presented as supplementary figures, which lack figure legends and are supported by only very brief captions. So, once again, I have been forced to make my best guess as to what these figures show and how it was computed.

Supplementary Figure 6a plots the "correlation" against "distance of CpG" ($0 - 1000$), which I believe are the genome-wide co-methylation estimates over the range $0 - 1000$ bp. Based on the "$n = 18,936,995$" quoted in the header of this figure, which I interpret as the number of pairs used in this plot of "correlation" against "istance of CpG" ($0 - 1000$), I believe that what is shown are the correlations of pairs of CpG $\beta$-values where the pairs are separated by a common distance and have no intervening CpGs. If all pairs were used this $n$ would be much higher since there are more than twenty million CpGs on one strand of the haploid human genome. To emphasise, the correlation shown by \cite{Li:2010fb} is not the same as the autocorrelation computed in \cite{Cokus:2008fc} since the autocorrelation uses all pairs of CpGs, regardless of the number of intervening CpGs.

The genome-wide analysis correlation plot contains a periodicity of approximately $170$ nucleotides, very similar to the $167$ nucleotide period identified in _Arabidopsis_ by \cite{Cokus:2008fc} (highlighted in Supplementary Figure 7). This is further evidence that nucleosome spacing effects the strength of co-methylation. However, \cite{Li:2010fb} found no evidence of higher frequency periods, such as the $10$ nucleotide period reported by \cite{Cokus:2008fc} and \cite{Chodavarapu:2010iqa}.

#### \cite{Lyko:2010dr}

In their study of differential methylation in queen and worker honey bees (_Apis mellifera_), \cite{Lyko:2010dr} perform an autocorrelation analysis of CpG methylation (Supplementary Figure 7). They report that the "correlation of methylation status of neighboring CpGs increases sharply between 1 bp and 20 bp, then drops rapidly between 40 bp and 100 bp, and then slowly fades away". This is the first report of an organism in which the correlation of methylation levels first increases before decreasing (__TODO: Is this the "first and only" such report? Check my data any such increases__). __TODO: Is this the fastest decay of correlation in any species?__ However, I am concerned that the maximum autocorrelation is less $0.015$, which doesn't seem to make sense and is very close to zero. This may simply be due to a mislabelled or "normalised" value being plotted as the "correlation". (__TODO: I have emailed the senior author, Ryszard Maleszka, to try to resolve this concern__).

Additionally, \cite{Lyko:2010dr} identify a $3$ nucleotide periodicity in the autocorrelation of CpG methylation and provide evidence that this "reflect[s] a preferential methylation of CpGs occupying the first and second position of the arginine codons", which is coded for by the DNA triplets CGT, CGC, CGA, CGG, AGA, and AGG (__CITE__). They do not detect a $10$ nucleotide periodicity, which the authors claim is due to "a much lower density of modified CpGs in this species [_Apis mellifera_]" than in human or _Arabidopsis_ genomes.

Since the authors specifically describe their analysis as an "autocorrelation", I believe it uses all pairs of CpGs regardless of the number of intervening CpGs.

#### \cite{Lacey:2013iy}

\cite{Lacey:2013iy} fit a Gaussian variogram model to the $\beta$-values at CpGs obtained via reduced representation bisulfite-sequencing (RRBS) of chromosome 11 from a normal myotube cell line (MTCTL2). The empirical and fitted variograms flatten out for CpGs separated by around $3000$ bp (Figure 1d) and so they conclude that the methylation level of CpGs "show a strong correlation for sites in close proximity, decaying to near independence at distances beyond 3000 bp".

Caution is warranted in extrapolating these results to statements about genome-wide co-methylation. Firstly, this analysis uses only approximately $60,000$ CpGs[^ncpgs], far fewer than the $25$ million CpGs (__TODO: Check number__) in the haploid human genome. Secondly, these data are from a single sample from a single tissue, which cannot capture the sample-to-sample nor tissue-to-tissue variability in co-methylation. Finally, RRBS targets CpG dense regions of the genome with low levels of methylation and so these data do not capture co-methylation outside of these regions, where the majority of the CpGs are found and which are highly methylated.

[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable RRBS libraries (__TODO: Cite and link to appendix__).

\subsection{Within-fragment co-methylation analyses}
\label{sec:previous_within-fragment_analyses}

Using bisulfite-sequencing technologies we can analyse co-methylation at the level of individual DNA molecules. I call this within-fragment co-methylation, where fragment refers to a fragment of DNA. More precisely, within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment. It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$. For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure. We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.

Because every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads. That is, by studying the methylation patterns at m-tuples. Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.

[^chimera]: Here I ignore the possibility of chimeric reads that may be artificially produced by DNA fragments ligating to one another during the library preparation or sequencing process (__TODO: reference__).

If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci. However, the most commonly used technology for high-throughput bisulfite-sequencing, which uses Illumina's  sequencing technology, can only generate reads that span approximately $200-250$ bp at a maximum. Assembling these reads into longer contigs will not solve this problem since this cannot guarantee that the assembled reads actually originated from the same DNA fragment.

Long-read sequencing technologies, such as Pacific Biosciences (__CITE__), Roche 454 (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lesser throughput of these technologies and the reliability with which they can identify methylcytosines create new complications (__CITE__). An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo](__CITE__).

[^moleculo]: My understanding is that Illumina's TruSeq Synthetic Long-Read technique cannot be directly applied to bisulfite-sequencing for a number of technique reasons. These include the fragility of large DNA fragments treated with sodium bisulfite (__TODO: Discuss with someone from AGRF__)

To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200$ nucleotides. There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels. Here I review this previously published work.

#### \cite{Cokus:2008fc}

As already discussed in Section \ref{sec:Cokus}, \cite{Cokus:2008fc} identify a $10$ nucleotide periodicity in the autocorrelation of $\beta$-values for CHH methylation from a whole-genome bisulfite-sequencing experiment of _Arabidopsis Thaliana_. They find this same periodicity "when individual reads [were] examined directly".

The sequencing technology at the time of this publication produced very short reads, on average only $31$ nucleotides of usable sequence. For the results shown in Supplementary Figure 14 of \cite{Cokus:2008fc}, they first identify all reads containing multiple CHH loci. Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated. This is estimated by simply counting the number of times this occurs and dividing by the number of pairs of CHH loci that occur within the same read[^cokus_clarification]. They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes. \cite{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci. This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of Normality, and a running average of these averages (). There is an approximate $10$ nucleotide periodicity evident in this plot, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.

[^cokus_clarification]: Actually, it's not clear whether they consider all pairs of CHH loci or only the first pair of CHH loci in a read since a read may contain multiple pairs. The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci. Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.

I consider these results to use a restricted definition of within-fragment "co-methylation". These results tells us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment. Specifically, \cite{Cokus:2008fc} only estimate $Pr(Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $Z_{h, i}$ and $Z_{h, i'}$ are two CHH loci on the same fragment $(i < i')$. From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y)$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$. It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$. While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{subsec:wf_cometh} that it is more useful and simpler to  estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.

Furthermore, by aggregating all pairs of CHH loci separated by a common distance and then estimating these probabilities, there is an implicit assumption that all such pairs on the same chromosome have the same conditional probability, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$. There is good reason to suspect that this is not true and that there exists significant pair-to-pair variability. For example, we know for CpGs that those within CpG islands have a vastly different probability of being methylated than do those outside of islands; it is reasonable to suspect that this is also true for CHH methylation in different regions of the _Arabidopsis_ genome and hence for "co-methylation" of these same loci.

#### \cite{Lacey:2009by}

\cite{Lacey:2009by} develop stochastic models of the replication of DNA methylation during mitosis (cell division) for a single double-stranded DNA molecule in humans. The most complex model, called the _neighbouring sites model_, uses a one-step Markov random field where the methylation state of one loci, $Z_{h, i}$, depends on that of its neighbours either side, $Z_{h, i - 1}$ and $Z_{h, i + 1}$. The physical distances between these neighbours, the $IPD$s, are not considered.

The neighbouring sites model requires an analysis of within-fragment co-methylation in order to estimate the transition probabilities. They estimate these parameters from two small regions, each approximately $200$ nucleotides in length and containing 13 and 14 CpGs, from 10 human ovarian carcinoma samples and a pool of somatic controls. These regions were sequenced using hairpin-bisulfite PCR, an assay that measures DNA methylation on both strands of a double-stranded DNA molecule[^hairpin_bs]. For each region, $12-16$ clones were sequenced per sample[^clones]

[^hairpin_bs]: This is unlike traditional bisulfite-sequencing, which only measures methylation along a single DNA strand.

[^clones]: Essentially, $12-16$ reads were generated per sample.

For each of these regions, \cite{Lacey:2009by} found that the methylation level of a CpG strongly depends on that of its neighbours. In other words, they found significant within-fragment co-methylation. They also found that the overall level of methylation of the CpGs in a region influenced the strength of the within-fragment co-methylation.

In summary, this analysis uses high resolution measurements from a moderate number of clones from a moderate number of samples for a miniscule fraction of the human genome. While the methods and models are quite interesting, the data are so focused that the results obtained from these analyses cannot represent a genome-wide picture of within-fragment co-methylation.

\subsection{Other related work}
\label{sec:previous_other_related_work}

Other definitions of co-methylation and related work:

* Lister's mCG correlations
* co-methylation across samples
* "vertical" correlations of \cite{Capra:2014uh}

#### \cite{Hebestreit:2013ko}

The `BiSeq` R/Bioconductor software uses a semivariogram to estimate the spatial correlation between test statistics of differential methylation at two methylation loci \cite{Hebestreit:2013ko}. This is not the same as looking for correlations between the $\beta$-values but rather looking for correlations between the test statistics, which themselves are based on _differences_ of $\beta$-values.

#### \cite{Liu:dy}

The focus of this study on identify clusters of correlated CpGs and relating methylation changes at these clusters to SNP genotypes. The authors compute the correlation of pairs of $\beta$-values across samples, which is analogous to estimating the linkage disequilibrium between a pair of single nucleotide polymorphisms (SNPs) across a set of samples. This is different to computing autocorrelations of $\beta$-values, which are computed within, rather than between, samples. An autocorrelation analysis implicitly assumes stationarity along the genome whereas this analysis assumes stationarity across samples (__TODO: Discuss with Terry__).

The discovery cohort contains 247 human samples typed on the Illumina HumanMethylation450 array. After normalisation and other pre-processing steps, they identify the top-$25\%$ most variably methylated CpG sites (vCpGs) in the 247 samples. They then compute Pearson correlations of methylation levels for all pairs of vCpGs across the samples.

From this they conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set". However, they also note that "methylation codependency (co-methylation) itself varies across the genome", as is to be expected.

The method I developed, which is described in Section \ref{subsec:wf_cometh}, is superior because of __TODO__.

>>>>

In this chapter I describe two measures of within-sample co-methylation, _within-fragment co-methylation_ and _correlation of $\beta$-values_. Using both these measures I explore how co-methylation is affected by the distance between methylation loci, the density of methylation loci and the genomic context of the methylation loci. I apply these methods to more than fifty whole-genome bisulfite-sequencing libraries to study the between-sample variability of co-methylation.

@@
